---
title: "Bisection Method for Finding Roots: Complete Explanation"
author: Sulagna Roy
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    highlight: tango
  pdf_document:
    toc: true
    number_sections: true
    keep_tex: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
rm(list = ls())
```

---

# Introduction

This document provides a **complete explanation** of the Bisection Method for finding roots of functions. The bisection method is a fundamental numerical technique used to locate zeros of functions by repeatedly dividing an interval in half and selecting the subinterval where the function changes sign.

**Key Concept:** If a continuous function f(x) has opposite signs at two points, then by the Intermediate Value Theorem, there must be at least one root between those points.

---

# Problem Setup

## The Functions

```{r define_functions}
# Define the objective function
f <- function(x) (x-2)^2 * (x-8)^2

# Define the derivative (for root finding)
df <- function(x) 4*(x-2)*(x-5)*(x-8)
```

## Mathematical Background

**Objective Function:**
$$f(x) = (x-2)^2(x-8)^2$$

This is a polynomial function with:
- Roots at x = 2 (double root) and x = 8 (double root)
- Minimum values at these roots
- Always non-negative: f(x) ≥ 0 for all x

**Derivative (Rate of Change):**
$$f'(x) = 4(x-2)(x-5)(x-8)$$

This derivative has:
- Roots at x = 2, x = 5, and x = 8
- Critical points (local extrema) at these values
- Sign changes at these points

## Why We Use the Derivative

The problem asks to **find the minimum of f(x) using the bisection method**. Instead of finding where f(x) = 0, we:

- Use the **derivative** f'(x) = 0 to find critical points
- The derivative changes sign at the minimum
- Apply bisection to find where f'(x) = 0

This is an application of **numerical root-finding** to solve the optimization problem.

---

# Initial Parameters

```{r initial_parameters}
a <- 6      # Left endpoint of interval
b <- 10     # Right endpoint of interval
e <- 0.01   # Tolerance (stopping criterion)
```

## Parameter Explanation

| Parameter | Value | Meaning |
|-----------|-------|---------|
| **a** | 6 | Left endpoint of search interval [6, 10] |
| **b** | 10 | Right endpoint of search interval [6, 10] |
| **e** | 0.01 | Tolerance/precision (interval width threshold) |

## Why These Values?

**Interval [6, 10]:**

- The derivative has a critical point at x = 5 and another at x = 8
- In the interval [6, 10], we expect to find a critical point near x = 8
- At x = 6: f'(6) = 4(4)(1)(−2) = −32 < 0 (function decreasing)
- At x = 10: f'(10) = 4(8)(5)(2) = 320 > 0 (function increasing)
- Sign change detected! Root must be between 6 and 10

**Tolerance e = 0.01:**

- Algorithm stops when interval width (b − a) ≤ 0.01
- This provides reasonable precision for finding the root
- Smaller e → more iterations, better precision
- Larger e → fewer iterations, less precision

---

# Bisection Algorithm

## Complete Code

```{r bisection_algorithm}
a <- 6
b <- 10
e <- 0.01

# Track iterations for visualization
iterations <- 0
history <- data.frame(
  Iteration = numeric(),
  a = numeric(),
  b = numeric(),
  m = numeric(),
  df_m = numeric(),
  width = numeric()
)

while ((b-a) > e) {
  iterations <- iterations + 1
  
  # Step 1: Find midpoint
  m <- (a+b)/2
  
  # Step 2: Evaluate derivative at midpoint
  df_m <- df(m)
  
  # Step 3: Bisection criterion
  if (df_m < 0) {
    # Derivative is negative (function decreasing)
    # Root is to the right of m
    a <- m
  } else {
    # Derivative is positive (function increasing)
    # Root is to the left of m
    b <- m
  }
  
  # Store iteration history
  history <- rbind(history, data.frame(
    Iteration = iterations,
    a = a,
    b = b,
    m = m,
    df_m = df_m,
    width = b - a
  ))
}

# Final estimate
x <- (a+b)/2

cat("Algorithm Convergence Summary:\n")
cat("===============================\n")
cat("Number of iterations:", iterations, "\n")
cat("Final interval width:", b-a, "\n")
cat("Tolerance threshold:", e, "\n")
```

---

# Algorithm Explanation: Step by Step

## Step 1: Find the Midpoint

```{r step1_explanation, eval=FALSE}
m <- (a+b)/2
```

**What it does:**
- Computes the midpoint m of the interval [a, b]
- m = (a + b) / 2
- This divides the interval into two equal halves

**Example:**
- If a = 6 and b = 10
- Then m = (6 + 10) / 2 = 8

---

## Step 2: Evaluate the Derivative at the Midpoint

```{r step2_explanation, eval=FALSE}
df_m <- df(m)
```

**What it does:**
- Calculates f'(m) at the midpoint
- This tells us the sign of the derivative at m

**Interpretation:**
- **f'(m) < 0**: Function is decreasing at m → Root is to the right (larger x)
- **f'(m) > 0**: Function is increasing at m → Root is to the left (smaller x)
- **f'(m) = 0**: Found the exact root! (unlikely with finite precision)

---

## Step 3: Update the Interval (Bisection Criterion)

```{r step3_explanation, eval=FALSE}
if (df_m < 0) {
  a <- m      # Move left endpoint to midpoint
} else {
  b <- m      # Move right endpoint to midpoint
}
```

**Logic:**

- **If f'(m) < 0** (derivative negative, function decreasing):
  - Root where derivative becomes zero is to the RIGHT of m
  - Keep everything to the right: Set a ← m
  - New interval: [m, b] (right half)
  
- **If f'(m) ≥ 0** (derivative positive/zero, function increasing):
  - Root where derivative becomes zero is to the LEFT of m
  - Keep everything to the left: Set b ← m
  - New interval: [a, m] (left half)

**Effect:** Interval shrinks by half each iteration

---

## Step 4: Check Stopping Criterion

```{r step4_explanation, eval=FALSE}
while ((b-a) > e) {
  # Continue loop
}
```

**Condition:**
- Loop continues as long as interval width > tolerance
- When (b − a) ≤ 0.01, algorithm terminates

**Precision Trade-off:**
- Smaller tolerance → More iterations, higher precision
- Larger tolerance → Fewer iterations, lower precision

---

# Iteration History

```{r show_history}
cat("\nDetailed Iteration History:\n")
cat("============================\n\n")
print(round(history, 6))
```

## What the History Shows

**Column Explanations:**

| Column | Meaning |
|--------|---------|
| **Iteration** | Which iteration (1st, 2nd, 3rd, ...) |
| **a** | Left endpoint after this iteration |
| **b** | Right endpoint after this iteration |
| **m** | Midpoint tested in this iteration |
| **df_m** | Derivative value at midpoint f'(m) |
| **width** | New interval width (b − a) |

**Key Observations:**

1. **Interval shrinks geometrically**: width ≈ width / 2 each iteration
2. **Sign change persists**: f'(a) and f'(b) always have opposite signs
3. **Convergence is guaranteed**: Width → 0 as iterations increase
4. **Linear convergence rate**: Each iteration eliminates half the remaining interval

---

# Final Results

```{r final_results}
cat("\n===============================\n")
cat("FINAL RESULTS\n")
cat("===============================\n\n")

x <- (a+b)/2

cat("Root location (x where f'(x) ≈ 0):\n")
cat("Minimum x =", x, "\n\n")

cat("Function value at root:\n")
cat("Minimum f(x) =", f(x), "\n\n")

cat("Verification:\n")
cat("f'(x) ≈", df(x), "\n")
cat("Derivative is very close to zero ✓\n")
```

## Interpretation

**Minimum x ≈ 7.99**
- The function f(x) = (x−2)²(x−8)² has a minimum near x = 8
- The derivative f'(x) = 4(x−2)(x−5)(x−8) equals zero at x = 8
- Algorithm found the root with tolerance 0.01

**Minimum f(x) ≈ 0**
- The function value at the minimum is approximately 0
- This makes sense: (7.99−2)²(7.99−8)² ≈ 0
- The minimum of f(x) is actually 0 (achieved at x = 2 and x = 8)

---

# Visualization

## Derivative Function Plot

```{r plot_derivative, fig.width=10, fig.height=6}
x_vals <- seq(0, 12, length = 500)
df_vals <- df(x_vals)

plot(x_vals, df_vals, type = 'l', lwd = 2, col = 'blue',
     main = "Derivative f'(x) = 4(x-2)(x-5)(x-8)",
     xlab = "x", ylab = "f'(x)",
     ylim = c(-50, 100))

# Add horizontal line at y = 0
abline(h = 0, col = 'black', lty = 2, lwd = 1)

# Mark the roots of the derivative
roots <- c(2, 5, 8)
abline(v = roots, col = 'red', lty = 3, lwd = 1.5, alpha = 0.5)
points(roots, rep(0, 3), pch = 16, col = 'red', cex = 2)

# Highlight search interval [6, 10]
rect(6, -50, 10, 100, border = 'green', lwd = 2, lty = 2)
text(8, 80, 'Search Interval [6,10]', col = 'green', cex = 1.2, font = 2)

# Mark final solution
final_x <- x
final_df <- df(final_x)
points(final_x, final_df, pch = 17, col = 'purple', cex = 2.5)
text(final_x, final_df + 10, paste('Solution:\nx ≈', round(final_x, 4)), 
     col = 'purple', cex = 1, font = 2)

legend('topleft', 
       c('f\'(x)', 'Roots of f\'(x)', 'Search interval', 'Found root'),
       col = c('blue', 'red', 'green', 'purple'),
       lty = c(1, 3, 2, NA),
       pch = c(NA, 16, NA, 17),
       lwd = c(2, 1.5, 2, NA),
       cex = 1)
```

## Objective Function Plot

```{r plot_objective, fig.width=10, fig.height=6}
f_vals <- f(x_vals)

plot(x_vals, f_vals, type = 'l', lwd = 2, col = 'darkgreen',
     main = "Objective Function f(x) = (x-2)²(x-8)²",
     xlab = "x", ylab = "f(x)",
     ylim = c(-5, 200))

# Mark critical points (where f'(x) = 0)
critical_points <- c(2, 5, 8)
f_critical <- f(critical_points)
points(critical_points, f_critical, pch = 16, col = 'red', cex = 2.5)

# Add labels
labels <- c("Min\n(f=0)", "Local\nMax", "Min\n(f=0)")
for(i in 1:3) {
  text(critical_points[i], f_critical[i] + 15, labels[i], 
       col = 'red', cex = 1, font = 2)
}

# Highlight search interval [6, 10]
rect(6, -5, 10, 200, border = 'blue', lwd = 2, lty = 2)
text(8, 150, 'Search Interval [6,10]', col = 'blue', cex = 1.2, font = 2)

# Mark found minimum
final_f <- f(final_x)
points(final_x, final_f, pch = 17, col = 'purple', cex = 2.5)
text(final_x - 0.8, final_f + 20, paste('Found:\nx ≈', round(final_x, 4), 
                                         '\nf(x) ≈', round(final_f, 4)), 
     col = 'purple', cex = 1, font = 2)

legend('topright',
       c('f(x)', 'Critical points', 'Search interval', 'Found minimum'),
       col = c('darkgreen', 'red', 'blue', 'purple'),
       lty = c(1, NA, 2, NA),
       pch = c(NA, 16, NA, 17),
       lwd = c(2, NA, 2, NA),
       cex = 1)
```

## Convergence Plot

```{r plot_convergence, fig.width=10, fig.height=6}
# Plot interval width over iterations
plot(history$Iteration, history$width, type = 'b', 
     pch = 16, cex = 1.5, col = 'darkblue', lwd = 2,
     main = "Convergence: Interval Width vs. Iteration",
     xlab = "Iteration Number",
     ylab = "Interval Width (b - a)",
     ylim = c(0, 4.5))

# Add tolerance line
abline(h = e, col = 'red', lty = 2, lwd = 2)
text(max(history$Iteration) * 0.7, e + 0.2, 
     paste('Tolerance = ', e), col = 'red', font = 2, cex = 1.1)

# Add grid
grid(col = 'gray', lty = 3)

# Annotate convergence
for(i in 1:nrow(history)) {
  text(history$Iteration[i], history$width[i] + 0.2, 
       round(history$width[i], 3), cex = 0.8, pos = 3)
}

legend('topright',
       c('Interval width', 'Tolerance threshold'),
       col = c('darkblue', 'red'),
       lty = c(1, 2),
       lwd = c(2, 2),
       pch = c(16, NA),
       cex = 1)
```

---

# Convergence Analysis

## Linear Convergence Rate

```{r convergence_analysis}
# Calculate convergence rate
initial_width <- history$width[1]
final_width <- history$width[nrow(history)]
n_iter <- nrow(history)

cat("\nConvergence Analysis:\n")
cat("=======================\n")
cat("Initial interval width:", initial_width, "\n")
cat("Final interval width:", final_width, "\n")
cat("Number of iterations:", n_iter, "\n\n")

# Theoretical reduction factor per iteration
reduction_per_iter <- 0.5
theoretical_final <- initial_width * (reduction_per_iter ^ n_iter)

cat("Theoretical convergence (width = initial × 0.5^n):\n")
cat("Expected final width:", round(theoretical_final, 6), "\n")
cat("Actual final width:", round(final_width, 6), "\n\n")

# Number of iterations needed for tolerance
target_width <- e
iterations_needed <- log(initial_width / target_width) / log(2)
cat("Theoretical iterations needed for tolerance e =", e, ":\n")
cat("log₂(", initial_width, "/", target_width, ") =", round(iterations_needed, 2), "\n")
cat("Actual iterations used:", n_iter, "\n")
```

---

# Summary: Key Concepts

## What is the Bisection Method?

The **bisection method** is a numerical algorithm to find roots of continuous functions by:

1. **Starting** with an interval [a, b] where f(a) and f(b) have opposite signs
2. **Finding** the midpoint m = (a + b) / 2
3. **Replacing** one endpoint with m to maintain opposite signs
4. **Repeating** until the interval is sufficiently small

## Advantages

✓ **Simple**: Easy to understand and implement  
✓ **Guaranteed**: Always converges if f is continuous and has a sign change  
✓ **Robust**: Works even if f is non-differentiable or has complex behavior  
✓ **Predictable**: Linear convergence rate is known  

## Disadvantages

✗ **Slow**: Requires many iterations compared to faster methods  
✗ **Sign change required**: Needs opposite signs at endpoints  
✗ **One root at a time**: Only finds one root per interval  
✗ **Cannot find complex roots**: Only works for real-valued functions  

## When to Use Bisection

- When you have an interval with a guaranteed sign change
- When f is continuous but difficult to differentiate
- When you need a simple, reliable method
- When multiple roots exist (use bisection separately for each region)
- When you're unsure if faster methods will converge

## Comparison with Other Root-Finding Methods

| Method | Speed | Convergence | Requirements |
|--------|-------|------------|--------------|
| **Bisection** | Slow (linear) | Guaranteed | Sign change |
| **Newton-Raphson** | Fast (quadratic) | Local | Derivative needed |
| **Secant** | Medium | Superlinear | No derivative |
| **Regula Falsi** | Medium | Linear | Sign change |

---

# Conclusion

The **bisection method** is a fundamental numerical technique for finding roots of functions. In this example, we used it to find the critical point (minimum) of f(x) = (x−2)²(x−8)² by finding where its derivative equals zero.

**Key Results:**
- Found minimum at x ≈ **7.99** (very close to the true value 8)
- Function value at minimum: f(x) ≈ **0.00** (essentially zero)
- Required **`r nrow(history)` iterations** to achieve tolerance 0.01
- **Linear convergence**: Interval width halves each iteration

The algorithm demonstrates the power of simple iterative methods for solving practical numerical problems when exact analytical solutions are unavailable.

---

# References

- Chapra, S. C., & Canale, R. P. (2015). Numerical Methods for Engineers (7th ed.). McGraw-Hill.

- Burden, R. L., & Faires, J. D. (2015). Numerical Analysis (10th ed.). Cengage Learning.

- Quarteroni, A., Sacco, R., & Saleri, F. (2010). Numerical Mathematics (2nd ed.). Springer.

