---
title: "Golden Section Search Method for Univariate Optimization"
author: "Sulagna Roy"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: '3'
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    theme: flatly
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Overview

This document explains an implementation of the **golden section search**
method, a derivative-free technique for finding the minimizer of a
*unimodal* function over a closed interval $[a, b]$. Like Fibonacci search,
it works by shrinking a bracketing interval guaranteed to contain the
minimum, evaluating the function at two interior points and discarding the
sub-interval that cannot contain the minimizer. Unlike Fibonacci search, it
uses a single fixed ratio — the **golden ratio conjugate**, $\tau \approx
0.618$ — at every iteration, so the number of iterations does not need to be
precomputed; the loop simply runs until the bracket width is smaller than a
chosen tolerance.

# The Objective Function

```{r objective-function}
# Objective function
f <- function(x) {
  (x - 2)^2 * (x - 8)^2
}
```

The function is

$$
f(x) = (x-2)^2 (x-8)^2,
$$

algebraically the same shape as $\big[(x-2)(x-8)\big]^2$: it touches zero at
$x=2$ and $x=8$ and is positive elsewhere, giving **two local minima**
separated by a local maximum. On the search interval $[1, 5]$ used below,
only the minimum at $x=2$ lies inside the bracket, so that is the one found.

# The Golden Section Search Function

```{r golden-search-function}
# Golden Section Search
golden_search <- function(f, a, b, tol = 1e-4) {
  
  tau <- (sqrt(5) - 1) / 2  # 0.618...
  
  history <- data.frame(
    iter = 0,
    a = a,
    b = b
  )
  
  x1 <- b - tau * (b - a)
  x2 <- a + tau * (b - a)
  
  f1 <- f(x1)
  f2 <- f(x2)
  
  iter <- 0
  while ((b - a) > tol) {
    iter <- iter + 1
    
    if (f1 > f2) {
      
      a <- x1
      x1 <- x2
      f1 <- f2
      
      x2 <- a + tau * (b - a)
      f2 <- f(x2)
      
    } else {
      
      b <- x2
      x2 <- x1
      f2 <- f1
      
      x1 <- b - tau * (b - a)
      f1 <- f(x1)
    }
    history <- rbind(
      history,
      data.frame(iter = iter, a = a, b = b)
    )
  }
  
  list(
    xmin = (a + b) / 2,
    interval = c(a, b),
    history = history
  )
}
```

## Explanation, step by step

**1. The golden ratio conjugate**

```r
tau <- (sqrt(5) - 1) / 2  # 0.618...
```

$\tau = \frac{\sqrt5 - 1}{2} \approx 0.618$ is the reciprocal of the golden
ratio $\varphi$. It has the special self-similar property
$\tau^2 = 1 - \tau$, which is exactly what allows one interior point to be
reused between iterations without shifting position relative to the new
bracket — the same mechanism Fibonacci search approximates using ratios of
Fibonacci numbers, except here it is a single constant rather than a
changing sequence.

**2. Initializing the history and interior points**

```r
history <- data.frame(iter = 0, a = a, b = b)
x1 <- b - tau * (b - a)
x2 <- a + tau * (b - a)
f1 <- f(x1)
f2 <- f(x2)
```

`history` records the bracket at iteration 0 (the original interval) before
any shrinking happens. Two interior points are placed symmetrically inside
$[a,b]$:

$$
x_1 = b - \tau(b-a), \qquad x_2 = a + \tau(b-a),
$$

with $x_1 < x_2$. Both points are evaluated once up front (`f1`, `f2`).

**3. The main loop**

```r
iter <- 0
while ((b - a) > tol) {
  iter <- iter + 1
  if (f1 > f2) {
    a <- x1; x1 <- x2; f1 <- f2
    x2 <- a + tau * (b - a); f2 <- f(x2)
  } else {
    b <- x2; x2 <- x1; f2 <- f1
    x1 <- b - tau * (b - a); f1 <- f(x1)
  }
  history <- rbind(history, data.frame(iter = iter, a = a, b = b))
}
```

Unlike Fibonacci search's `for` loop over a precomputed number of steps,
this is a `while` loop that runs until the raw bracket width `b - a` drops
below `tol`. Each pass:

- **If $f(x_1) > f(x_2)$:** the minimum cannot lie left of $x_1$, so
  `a <- x1`. The old $x_2$ slides into the $x_1$ role (reusing `f2` as the
  new `f1`), and a fresh $x_2$ is computed with the same formula
  `a + tau * (b - a)` against the *new*, narrower interval, needing one new
  evaluation of `f`.
- **Otherwise:** symmetric — `b <- x2`, the old $x_1$ becomes the new $x_2$
  (reusing `f1` as `f2`), and a fresh $x_1$ is computed via
  `b - tau * (b - a)`, again one new evaluation.

Every iteration therefore costs exactly **one** new function evaluation,
the same efficiency Fibonacci search achieves, but here obtained from the
fixed-point property of $\tau$ rather than from a precomputed sequence. The
updated bracket is appended to `history` after each pass.

**4. Returning the result**

```r
list(
  xmin = (a + b) / 2,
  interval = c(a, b),
  history = history
)
```

When the loop exits, `b - a \le tol`, so the bracket is tiny. The function
returns the midpoint of that final bracket as `xmin`, the final `interval`
itself, and the full `history` of every iteration's bracket for inspection
or plotting.

# Running the Search

```{r run-search}
# Run Golden Section Search
result <- golden_search(f, a = 1, b = 5, tol = 1e-4)

print(result)
```

## Interpreting the result

On $[a,b] = [1,5]$ with `tol = 1e-4`:

- `xmin` $\approx 1.999998$ — essentially exact, since the true minimizer is
  $x=2$.
- `interval` $\approx [1.999967,\ 2.000029]$ — a final bracket of width about
  $6\times 10^{-5}$, comfortably under the $10^{-4}$ tolerance.
- `history` has `r nrow(result$history)` rows (iterations $0$ through
  $23$), meaning the `while` loop executed 23 times before `b - a` fell
  below `tol`. As with Fibonacci search, $a$ rises and $b$ falls
  monotonically, but here the shrink factor per step is the constant $\tau
  \approx 0.618$, so the bracket width shrinks geometrically rather than by
  a step-dependent Fibonacci ratio.

```{r show-history}
h <- result$history
h
```

# Visualizing the Convergence

```{r plot-convergence, fig.width=7, fig.height=5}
u <- seq(0, 6, length = 500)
plot(u, f(u), type = 'l',
     main = "Golden Section Search: Shrinking Bracket over f(x) = (x-2)^2(x-8)^2",
     xlab = "x", ylab = "f(x)")

for (i in 1:nrow(h)) {
  abline(v = h[i, 2], col = 2)  # left endpoint a, in red
  abline(v = h[i, 3], col = 3)  # right endpoint b, in green
}
```

## Reading the plot

The black curve is $f(x) = (x-2)^2(x-8)^2$ over $x \in [0,6]$, dipping to
zero at $x=2$ and beginning to rise again toward the second (unreached)
minimum near $x=8$, just visible climbing at the right edge of the plot.

- **Red lines** are the left endpoint $a$ (column 2 of `h`) at each
  iteration.
- **Green lines** are the right endpoint $b$ (column 3 of `h`) at each
  iteration.

The first red/green pair spans the full starting interval $[1,5]$; with each
successive iteration a new red line sits a little further right and a new
green line a little further left, so the bundle of lines visibly narrows
around $x=2$. By the final iterations the red and green lines are packed
together indistinguishably at the bottom of the left dip, matching the
converged `interval` reported above.

# Summary

This script implements golden section search:

1. It uses the single constant $\tau = (\sqrt5-1)/2$ to place two interior
   points inside the bracket at every iteration.
2. Its self-similarity property lets one evaluation be reused each step, so
   only one new evaluation of `f` is needed per iteration.
3. A `while` loop shrinks the bracket geometrically by factor $\tau$ until
   its width is below `tol`, with no need to precompute the iteration count.
4. It records the bracket history for plotting and returns the midpoint of
   the final tiny bracket as the estimated minimizer.

For $f(x) = (x-2)^2(x-8)^2$ on $[1,5]$, the method converges to
$x \approx 2.000000$ in 23 iterations, matching the true local minimizer to
well within the requested tolerance.
