---
title: "Golden Section Search Applied to a Lognormal-Type Objective"
author: "Sulagna Roy"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    theme: flatly
    df_print: paged
  pdf_document:
    toc: true
    toc_depth: '3'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Overview

This document explains a second application of the **golden section search**
method, this time minimizing a function built from the **lognormal
distribution's density shape** rather than a simple polynomial. The
algorithm itself is identical to the earlier golden-section example — the
only change is the objective function `f` and the search interval passed to
it. This document focuses the explanation on what makes this new objective
different and how that changes the numerical behavior of the search.

# The Objective Function

```{r objective-function}
# Objective function
f <- function(x) {
  -(1/x*sqrt(2*pi))*exp(-0.5*(log(x)*log(x)))
}
```

## What this function computes

The (standard) lognormal probability density function is

$$
\phi_{\text{LN}}(x) = \frac{1}{x\sqrt{2\pi}} \exp\!\left(-\frac{(\ln x)^2}{2}\right), \qquad x > 0,
$$

which is exactly the *positive* of the expression being computed here,
before the leading minus sign. So `f(x)` in this script is

$$
f(x) = -\phi_{\text{LN}}(x),
$$

i.e. the **negative** of the standard lognormal density. This sign flip is
a standard trick: golden section search (like the earlier examples) is
written to find a **minimum**, but here the real goal is to find the
$x$ that **maximizes** the lognormal density — its mode. Minimizing $-\phi_{\text{LN}}(x)$
is mathematically equivalent to maximizing $\phi_{\text{LN}}(x)$, so no
change to the search algorithm itself is needed.

## Domain consideration

Because of the $1/x$ term and $\ln x$, this function is only defined for
$x > 0$; at $x = 0$ it is undefined (division by zero and $\ln 0$).
This is why the search below is run starting from `a = 0.01` rather than
`a = 0`, keeping the search safely inside the function's valid domain
while still starting very close to zero.

## Shape of the function

Unlike the earlier quartic examples, which had two symmetric-looking dips,
$-\phi_{\text{LN}}(x)$ has a **single** minimum. The lognormal density
rises steeply from 0 near $x=0^+$, peaks at its mode, and then decays
slowly toward 0 as $x \to \infty$ — so its negative has a single sharp dip
at that same mode and flattens out toward 0 from below as $x$ grows. This
means $f$ is unimodal on any interval containing that one dip, exactly the
condition golden section search requires.

The mode of a standard lognormal distribution (i.e. $\ln X \sim N(0,1)$)
has the closed-form value $x^\* = e^{-\sigma^2} = e^{-1} \approx 0.367879$,
which will serve as a useful check against the numerical answer below.

# The Golden Section Search Function

```{r golden-search-function}
# Golden Section Search
golden_search <- function(f, a, b, tol = 1e-4) {
  
  tau <- (sqrt(5) - 1) / 2 # 0.618... 
  
  history <- data.frame(
    iter = 0,
    a = a,
    b = b
  )
  
  x1 <- b - tau * (b - a)
  x2 <- a + tau * (b - a)
  
  f1 <- f(x1)
  f2 <- f(x2)
  
  iter <- 0
  while ((b - a) > tol) {
    iter <- iter + 1
    
    if (f1 > f2) {
      
      a <- x1
      x1 <- x2
      f1 <- f2
      
      x2 <- a + tau * (b - a)
      f2 <- f(x2)
      
    } else {
      
      b <- x2
      x2 <- x1
      f2 <- f1
      
      x1 <- b - tau * (b - a)
      f1 <- f(x1)
    }
    history <- rbind(
      history,
      data.frame(iter = iter, a = a, b = b)
    )
  }
  
  list(
    xmin = (a + b) / 2,
    interval = c(a, b),
    history = history
  )
}
```

## Explanation, step by step

This is the same golden section search routine used previously, so the
logic is summarized here rather than re-derived from scratch:

**1. Golden ratio conjugate.** `tau <- (sqrt(5) - 1) / 2` sets
$\tau \approx 0.618$, the constant satisfying the self-similarity property
$\tau^2 = 1-\tau$. This is what allows one interior evaluation point to be
recycled at every iteration.

**2. Initial bracketing.** Two interior points are placed inside $[a,b]$:

$$
x_1 = b - \tau(b-a), \qquad x_2 = a + \tau(b-a),
$$

and `f` — here, the negative lognormal density — is evaluated once at each
to get `f1` and `f2`.

**3. Iterative shrinking (`while` loop).** As long as the raw bracket width
`b - a` exceeds `tol`:

- **If `f1 > f2`:** the (negative-density) value at the left test point is
  larger, so under unimodality the mode-minimizer cannot be left of `x1`;
  the left endpoint advances, `a <- x1`, the old `x2` is recycled as the
  new `x1`, and a single new `x2` is computed and evaluated.
- **Otherwise:** the right endpoint contracts, `b <- x2`, the old `x1`
  becomes the new `x2`, and a single new `x1` is computed and evaluated.

Exactly one new call to `f` happens per iteration, and the updated bracket
is appended to `history`.

**4. Return value.** Once `b - a <= tol`, the loop stops and the function
returns the midpoint `(a + b) / 2` as `xmin`, the tiny final `interval`,
and the complete `history` of every bracket along the way.

The only place this objective function's different shape matters
numerically is in *how many* iterations are needed and *where* the
interval starts, both of which are visible in the results below.

# Running the Search

```{r run-search}
# Run Golden Section Search
result <- golden_search(f, a = 0.01, b = 10, tol = 1e-4)

print(result)
```

## Interpreting the result

This time the search interval is $[a,b] = [0.01,\ 10]$ — much wider than
the $[1,5]$ interval used for the polynomial examples, because the
lognormal density's interesting behavior (its rise, peak, and long decay)
spans a wider range and the domain restriction $x>0$ makes starting near 0
natural.

- `xmin` $\approx$ `r round(result$xmin, 7)`, extremely close to the exact
  closed-form mode $e^{-1} \approx 0.367879$ derived above — confirming the
  search correctly located the peak of the lognormal density (since it was
  minimizing the *negative* density).
- `interval` $\approx$ `r paste0("[", paste(round(result$interval, 7), collapse=", "), "]")`,
  a final bracket narrower than the `tol = 1e-4` requirement.
- `history` has `r nrow(result$history)` rows (iterations $0$ through
  `r max(result$history$iter)`), noticeably **more iterations than the
  earlier $[1,5]$ examples** (which needed 23). This is expected: golden
  section search shrinks the *absolute* width of the bracket by a factor of
  $\tau$ each step, and the starting width here is $10 - 0.01 = 9.99$,
  nearly double the width of $5-1=4$ used previously — a wider starting
  interval simply takes more $\tau$-multiplications to shrink down to the
  same absolute tolerance.

```{r show-history}
h <- result$history
h
```

Looking at the printed history, notice how for the first several
iterations only the *left* endpoint `a` stays fixed at `0.01` while `b`
repeatedly contracts — this reflects the steep rise of $\phi_{\text{LN}}$
just to the right of 0 pulling the search rightward quickly, before the
bracket narrows enough for both endpoints to start moving on alternating
iterations.

# Visualizing the Convergence

```{r plot-convergence, fig.width=7, fig.height=5}
u <- seq(0, 6, length = 500)
plot(u, f(u), type = 'l',
     main = "Golden Section Search on the Negative Lognormal Density",
     xlab = "x", ylab = "f(x) = -lognormal density")

for (i in 1:nrow(h)) {
  abline(v = h[i, 2], col = 2)  # left endpoint a, in red
  abline(v = h[i, 3], col = 3)  # right endpoint b, in green
}
```

## Reading the plot

**Note on the plotting range:** the curve is drawn over `u = seq(0, 6,
length = 500)`, i.e. only $x \in [0,6]$, even though the search itself ran
on the much wider bracket $[0.01, 10]$. This is a minor mismatch in the
original script — the plot window happens to still comfortably contain the
minimizer near $x \approx 0.368$ and the final, fully-converged bracket
lines, but it does **not** show the early iterations' bracket endpoints
that briefly sit outside $[0,6]$ (for example the very first `b = 10`).
Those early vertical lines are simply not drawn because they fall outside
the visible x-axis range; this does not affect the correctness of the
numerical result, only what is visible in this particular plot.

Within the visible window:

- The black curve shows $f(x)=-\phi_{\text{LN}}(x)$ plunging from
  near 0 at $x=0^+$ down to a sharp minimum around $x\approx 0.37$, then
  rising smoothly back toward 0 as $x$ increases — the mirror image of the
  lognormal density's familiar right-skewed hump.
- **Red lines** mark left endpoints `a`, **green lines** mark right
  endpoints `b`, drawn for every iteration in `history` whose value lies
  inside $[0,6]$.
- As with the earlier examples, the bundle of red and green lines
  visibly tightens around the minimum near $x\approx0.37$, with the
  final several iterations' lines essentially overlapping — visual
  confirmation that the bracket has collapsed onto the mode of the
  lognormal density.

# Summary

This example reuses the same golden section search engine as before,
demonstrating that the method generalizes immediately to any unimodal
objective, not just polynomials:

1. The objective here is the *negative* of the standard lognormal density,
   turning a maximization (find the density's mode) into the minimization
   golden section search is built for.
2. The function's domain restriction ($x>0$) is respected by starting the
   search bracket at `a = 0.01` instead of 0.
3. The wider starting interval ($[0.01, 10]$ vs. $[1,5]$ previously) means
   more iterations are needed to reach the same absolute tolerance, since
   each step shrinks the bracket by the same multiplicative factor $\tau$
   but starts from a larger absolute width.
4. The numerical result, `xmin` $\approx$ `r round(result$xmin, 6)`, matches
   the known closed-form mode of the standard lognormal distribution,
   $e^{-1}\approx 0.367879$, to within the requested tolerance — a useful
   sanity check that the search implementation is correct.
