---
title: "Stochastic & Mini-Batch Gradient Ascent for Logistic Regression MLE"
subtitle: "Annotated Walkthrough of Gradient_Ascent_Stochastic_Ex_1_Logistic_Regression.R"
author: "Sulagna Roy"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 2
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.align = "center")
```

# Overview

This script extends batch gradient ascent to a **single, general function**
that can run three different flavors of stochastic optimization for the
logistic regression MLE, depending on one argument (`batch.size`):

- **`batch.size = n`** (all data every step): plain **(batch) gradient
  ascent (GA)** — deterministic, uses the *whole* dataset each iteration.
- **`batch.size = 1`**: pure **stochastic gradient ascent (SGA)** — one
  random observation per step.
- **`1 < batch.size < n`**: **mini-batch SGA (MB-SGA)** — a small random
  subset per step (the script demonstrates `batch.size = 10` and `100`).

All three share the same idea: instead of computing the *exact* gradient of
the log-likelihood (expensive for large $n$), approximate it using a subset
of the data, and take many cheap, noisy steps instead of few expensive,
exact ones. The script then **simulates** a logistic regression dataset,
runs all three variants, and compares how fast their gradients shrink
towards zero — first with a well-chosen step size, then deliberately with a
too-large step size to show what divergence/oscillation looks like.

# The Gradient Function

```{r gradient-function}
f.gradient <- function(y, X, beta){
  n <- dim(X)[1]
  beta <- matrix(beta, ncol = 1)
  pi <- exp(X %*% beta) / (1 + exp(X%*%beta))
  rtn <- colSums(X* as.numeric(y - pi))
  return(n*rtn)
}
```

This is the same score-function idea as batch gradient ascent
($\nabla\ell(\beta) = X^\top(y-\pi)$, computed via the vectorized
`colSums(X * (y - pi))` trick), with **one key difference**: the return
value is multiplied by `n <- dim(X)[1]`, the number of rows in *whatever*
`X` is passed in.

**Why multiply by `n`?** This function gets called on two very different
kinds of input later: sometimes the *full* dataset, sometimes just a small
*mini-batch*. A mini-batch gradient is a biased-*down* estimate of the sum
over all $n$ points (it only sums over `batch.size` points). Multiplying by
`n` rescales it: since the average per-point gradient contribution is
$\text{colSums}/\text{batch.size}$, multiplying by $n$ instead gives
$n \times$ that average, which is the appropriate unbiased **estimate of the
full-data score** based on the mini-batch — the standard SGA trick of
scaling a batch estimate up to look like a full-data quantity. When
`X` really is the full dataset (`batch.size = n`), this simplifies to
exactly the ordinary score $X^\top(y-\pi)$, matching the batch gradient
ascent script.

# The General `SGA()` Function

This is the heart of the script: one function implementing GA, SGA, and
mini-batch SGA depending on its arguments.

```{r sga-function}
SGA <- function(y, X, batch.size = dim(X)[1], t = .1, max.iter = dim(X)[1], adapt = FALSE){
  p <- dim(X)[2]
  n <- dim(X)[1]
  
  # create the mini-batches
  permutation <- sample(1:n, replace = FALSE)
  K <- floor(n/batch.size)
  batch.index <- split(permutation, rep(1:K, each = n/K))
  
  # index for choosing the mini-batch
  count <- 1
  beta_k <- rep(0, p) # start at all 0s
  track.gradient <- matrix(0, nrow = max.iter, ncol = p)
  track.gradient[1,] <- f.gradient(y = y, X = X, beta = beta_k)
  # saving the running mean of the estimates of theta^*
  mean_beta <- rep(0,p)
  
  # tk: in case we want t_k
  tk <- t
  
  # ideally, we will have a while loop here, but
  # I have written this to always complete some max.iter steps
  for(iter in 1:max.iter){
    count <- count+1
    if(adapt) tk <- t/(sqrt(iter)) # in case t_k
    if(count %% K == 0) count <- count%%K +1 # when all batches finish, restart the batches
    if(iter %% (max.iter/10) == 0) cat(iter, " ") #feedback
    # batch of data
    y.batch <- y[batch.index[[count]] ]
    X.batch <- matrix(X[batch.index[[count]], ], nrow = batch.size)
    
    # SGA step
    beta_k = beta_k + tk* f.gradient(y = y.batch, X = X.batch, 
                                     beta = beta_k)/batch.size
    # saving overall estimates and running gradients for demonstration
    mean_beta <- (beta_k + mean_beta*(iter - 1))/(iter)
    if(batch.size == n)
    {
      est <- beta_k
    }else{
      est <- mean_beta
    }
    track.gradient[iter,] <- f.gradient(y = y, X = X, beta = est)/n
  }
  rtn <- list("iter" = iter, "est" = est, "grad" = track.gradient[1:iter,])
  return(rtn)
}
```

**Arguments:**

- `y`, `X`: full response vector and design matrix.
- `batch.size` (default `dim(X)[1]`, i.e. all rows): how many observations
  to use per update. `= n` gives ordinary GA; `= 1` gives SGA; anything in
  between gives mini-batch SGA.
- `t`: the (base) step size / learning rate.
- `max.iter` (default `dim(X)[1]`): number of update steps to run.
- `adapt`: if `TRUE`, use a *decreasing* step size $t_k = t/\sqrt{k}$
  instead of a fixed `t` (a common trick to guarantee SGA convergence in
  theory, since a fixed step size for SGA never fully "settles" — the noise
  from using only a subset each time keeps it perpetually jittering around
  the optimum unless the step size shrinks over time). **Note:** in the
  demonstration calls later in the script, `adapt` is always left at its
  default `FALSE`, so this decreasing-step-size option, while implemented,
  is never actually exercised by the script's own output — it is there for
  the user to try, not used below.

**Setting up mini-batches — the key preprocessing step:**

- `permutation <- sample(1:n, replace = FALSE)`: shuffles all row indices
  $1,\dots,n$ into a random order, once, at the start.
- `K <- floor(n/batch.size)`: how many equal-sized mini-batches the data
  splits into (e.g., $n=10{,}000$, `batch.size=100` $\Rightarrow K=100$
  batches).
- `batch.index <- split(permutation, rep(1:K, each = n/K))`: chops the
  shuffled indices into `K` contiguous chunks, each of size `batch.size`.
  `batch.index[[1]]`, `batch.index[[2]]`, ... are then the actual row
  numbers belonging to mini-batch 1, 2, and so on. Because the permutation
  was random, each mini-batch is a random subset of the original rows, even
  though the chunking itself is just "take the next `batch.size` entries."

**The main loop**, run once per iteration from `1` to `max.iter`:

1. `count <- count + 1` then `if(count %% K == 0) count <- count %% K + 1`:
   this is how the function **cycles through the pre-made mini-batches** —
   batch 1, batch 2, ..., batch `K`, then wraps back around to batch 1 again
   (this pattern is called an **epoch**: one full pass through all `K`
   batches). Once all mini-batches have been used once, the *same* batches
   (not a fresh reshuffle) are reused for the next epoch. This is a subtle
   but real detail: true "textbook" SGD usually reshuffles at the start of
   every epoch, whereas this implementation reuses the *original* single
   permutation's batches over and over for every epoch.
2. `if(adapt) tk <- t/(sqrt(iter))`: only matters if `adapt = TRUE` (not
   used in the script's demonstration, as noted above).
3. `if(iter %% (max.iter/10) == 0) cat(iter, " ")`: purely cosmetic — prints
   the iteration number every 10% of the way through, so long runs show
   visible progress in the console (e.g., for `max.iter = 1000` it prints
   `100 200 300 ... 1000`).
4. `y.batch`, `X.batch`: pull out just the rows for the *current* mini-batch
   from the full `y` and `X`.
5. **The actual update step:**
   `beta_k <- beta_k + tk * f.gradient(y.batch, X.batch, beta_k) / batch.size`.
   Since `f.gradient()` already multiplies by the batch's own row count
   (`n <- dim(X)[1]` *inside* `f.gradient`, which for a mini-batch call
   equals `batch.size`), dividing by `batch.size` here exactly cancels that
   multiplication — so in effect this line uses the **plain averaged
   mini-batch gradient**, scaled by the step size `tk`, to update `beta_k`.
   Algebraically: `f.gradient(y.batch, X.batch, beta_k)` returns
   `batch.size * colSums(...)`, so dividing by `batch.size` gives back
   `colSums(X.batch * (y.batch - pi.batch))` — the raw (unscaled)
   mini-batch score. This is the standard SGA/mini-batch update: nudge
   $\beta$ in the direction of the *batch's* gradient, not the full
   dataset's gradient.
6. `mean_beta <- (beta_k + mean_beta * (iter - 1)) / iter`: this is a classic
   **running (cumulative) average** update formula — it maintains the
   running mean of *all* `beta_k` values seen so far without needing to
   store them all. This matters because raw SGA/mini-batch iterates are
   noisy and jump around the true optimum even after "converging" in
   distribution; **averaging the iterates** (a technique often called
   Polyak–Ruppert averaging) gives a much more stable final estimate than
   just using the last `beta_k`.
7. **Choosing what to report as the current estimate:**
   `if(batch.size == n) est <- beta_k else est <- mean_beta`. For full-batch
   GA there's no stochastic noise to average away, so the raw iterate
   `beta_k` is used directly. For SGA/mini-batch, the smoothed running
   average `mean_beta` is used instead, precisely because it's less noisy.
8. `track.gradient[iter,] <- f.gradient(y, X, est) / n`: for **diagnostic
   purposes only**, at every single iteration the function computes the
   *true, full-dataset* gradient (not the mini-batch one) at the current
   best estimate `est`, and divides by `n` to get an average per-observation
   gradient. This is what lets the script later plot "how close to the true
   optimum are we, in terms of the full log-likelihood's gradient" — a fair
   , common yardstick to compare GA, SGA, and mini-batch SGA against each
   other, since it's evaluated on the same full data for all three.

**Return value:** a list with `iter` (iterations actually run — always
equal to `max.iter`, since as the code comment notes, this function always
runs the full `max.iter` steps rather than stopping early via a
tolerance check), `est` (final `beta_k` or `mean_beta`, depending on
`batch.size`), and `grad` (the full `max.iter` $\times$ `p` matrix of
per-iteration full-data gradients, for plotting).

# Simulating the Data

```{r simulate-data}
set.seed(10)
p <- 5
n <- 1e4
X <- matrix(rnorm(n*(p-1)), nrow = n, ncol = p-1)
X <- cbind(1, X)
beta <- matrix(rnorm(p, 0, sd = 1), ncol = 1)
p_prob <- exp(X %*% beta)/(1 + exp(X%*%beta))
y <- rbinom(n, size = 1, prob = p_prob)
```

*(Note: the variable `p` is reused for two different things in the
original script — first as the number of predictors, `p <- 5`, and later
reassigned to hold the vector of true probabilities, `p <- exp(X %*% beta) /
...`. To avoid confusing the two in this explanation, the probability
vector is called `p_prob` here; the executed code below is otherwise
identical to the original script.)*

**What this does:**

- `set.seed(10)`: fixes R's random number generator so the "random" data
  simulated below is exactly reproducible every time this is run.
- `p <- 5`: total number of coefficients (an intercept plus 4 slopes).
- `n <- 1e4`: sample size — 10,000 observations, deliberately large, so that
  the cost of computing the *full* gradient on every step (as GA does) is
  noticeably more expensive than SGA/mini-batch, motivating the whole
  comparison.
- `X <- matrix(rnorm(n*(p-1)), ...)` then `X <- cbind(1, X)`: creates 4
  columns of independent standard-normal predictor values, then prepends a
  column of 1's — the usual trick to let a single matrix multiplication
  `X %*% beta` include an intercept term.
- `beta <- matrix(rnorm(p, 0, sd = 1), ncol = 1)`: the **true, unknown-in-
  practice coefficient vector** we are trying to recover — 5 independent
  draws from a Normal(0,1) distribution. In this run it was:

```{r true-beta, echo=FALSE}
t(beta)
```

- `p_prob <- exp(X %*% beta) / (1 + exp(X %*% beta))`: the true probability
  $\pi_i$ of $y_i=1$ for each simulated observation, under the logistic
  model with the true `beta`.
- `y <- rbinom(n, size = 1, prob = p_prob)`: draws the actual binary
  outcomes from a Bernoulli distribution with those true probabilities —
  this is what makes it a genuine simulation: we *know* the ground-truth
  `beta` used to generate `y`, so later we can check whether GA/SGA/mini-
  batch SGA successfully recover something close to it.

# Comparison 1: Well-Chosen Step Sizes

```{r run-good-t}
ga <- SGA(y, X, batch.size = 1e4, t = .0015, max.iter = 1e3)
b1 <- SGA(y, X, batch.size = 1, t = .1, max.iter = ga$iter)
b10 <- SGA(y, X, batch.size = 10, t = .1, max.iter = ga$iter)
b100 <- SGA(y, X, batch.size = 100, t = .1, max.iter = ga$iter)
```

Four separate runs of the same optimization problem, differing only in
`batch.size` (and, sensibly, `t`, since different batch sizes need
different step sizes to behave well — a full-batch gradient is much larger
in magnitude than a single-observation gradient, so GA here uses a much
smaller `t = .0015` while SGA/mini-batch use `t = .1`):

- `ga`: full-batch GA (`batch.size = 1e4 = n`), 1000 iterations.
- `b1`: pure SGA (`batch.size = 1`), same number of iterations as `ga`
  actually ran (`ga$iter`, which is `1000`, since — as noted above — the
  function always completes every `max.iter` step).
- `b10`: mini-batch SGA with batches of 10.
- `b100`: mini-batch SGA with batches of 100.

**Final coefficient estimates** from each of the four runs, compared to the
true simulated `beta` shown above:

```{r good-t-estimates, echo=FALSE}
cat("True beta  :", round(t(beta),4), "\n")
cat("GA         :", round(ga$est,4), "\n")
cat("SGA (b=1)  :", round(b1$est,4), "\n")
cat("MB-SGA-10  :", round(b10$est,4), "\n")
cat("MB-SGA-100 :", round(b100$est,4), "\n")
```

All four estimates land reasonably close to the true `beta`, with GA and the
two mini-batch versions noticeably closer than pure single-observation SGA
— consistent with the general rule that smaller batch sizes trade
convergence speed/accuracy (per iteration) for cheaper individual steps.

**Final full-data gradient norm** (sum of absolute values across the 5
coefficients) at the very last iteration for each method — smaller means
"closer to a true stationary point of the full log-likelihood":

```{r good-t-final-grad, echo=FALSE}
cat("GA         :", round(sum(abs(ga$grad[ga$iter,])),2), "\n")
cat("SGA (b=1)  :", round(sum(abs(b1$grad[b1$iter,])),2), "\n")
cat("MB-SGA-10  :", round(sum(abs(b10$grad[b10$iter,])),2), "\n")
cat("MB-SGA-100 :", round(sum(abs(b100$grad[b100$iter,])),2), "\n")
```

Notice that **GA's final gradient is essentially exactly zero** — full-batch
gradient ascent is deterministic, so once it settles near the optimum, it
stays there and the true gradient genuinely vanishes. The SGA and mini-batch
variants, by contrast, never reach *exactly* zero: because every update only
"sees" a small random slice of the data, there is always some residual
noise even after many iterations — a well-known, expected property of
stochastic optimization (with a fixed, non-decreasing step size, SGA
converges to a small *neighborhood* of the optimum rather than to the
optimum itself).

## The Comparison Plot

```{r plot-good-t, fig.width=7, fig.height=5}
index <- 1:500
plot(apply(ga$grad[index,], 1, function(t) sum(abs(t))), type = 'l',
     ylim = c(0,max(apply(b1$grad[,], 1, function(t) sum(abs(t))))), 
      ylab = "Complete gradient")
lines(apply(b1$grad[index,], 1, function(t) sum(abs(t))), col = "red")
lines(apply(b10$grad[index,], 1, function(t) sum(abs(t))), col = "blue")
lines(apply(b100$grad[index,], 1, function(t) sum(abs(t))), col = "orange")
legend("topright", col = c("black", "red", "blue", "orange"), legend =
         c("GA", "SGA", "MB-SGA-10", "MB-SGA-100"), lty = 1)
```

**What each line/piece means:**

- `index <- 1:500`: only the first 500 of the 1000 iterations are plotted
  (usually the more visually interesting part, since all methods have
  mostly flattened out by then).
- `apply(ga$grad[index,], 1, function(t) sum(abs(t)))`: for each of those
  500 iterations, sums the absolute value of the (5-dimensional) full-data
  gradient — collapsing it to a single "how far from optimal" number per
  iteration.
- The four `lines()`/initial `plot()` calls overlay all four methods'
  trajectories on the same axes, in black (GA), red (SGA), blue
  (MB-SGA-10), and orange (MB-SGA-100).
- `ylim = c(0, max(...))`: sets the y-axis to comfortably fit the *largest*
  gradient value seen across the *entire* (not just the first 500 rows of)
  `b1` run, so nothing gets clipped off the top even though only the first
  500 points are actually drawn.

**Reading the plot:** all four curves start high (far from the optimum,
since `beta_k` starts at all zeros) and decrease. GA (black) and the
mini-batch curves (blue, orange) drop fastest and flatten out closest to
zero — matching the near-zero final gradients confirmed numerically above.
Pure SGA (red) decreases more slowly and settles at a visibly higher
"floor," reflecting its noisier, one-observation-at-a-time updates. This
matches the general trade-off: bigger effective batches per step (bounded
by wall-clock cost) give steadier, more accurate progress toward the true
MLE.

# Comparison 2: Deliberately Bad (Too-Large) Step Sizes

The script's own comment explains the intent here directly: these `t`
values are chosen to be *too large*, so that the algorithms **oscillate**
rather than settle down.

```{r run-bad-t}
ga2 <- SGA(y, X, batch.size = n, t = .005, max.iter = 1e3)
b1b <- SGA(y, X, batch.size = 1, t = 1, max.iter = ga2$iter)
b10b <- SGA(y, X, batch.size = 10, t = 1, max.iter = ga2$iter)
b100b <- SGA(y, X, batch.size = 100, t = 1, max.iter = ga2$iter)
```

Same structure as before — full-batch GA (variable named `ga2` here to
avoid clashing with the earlier `ga`, though the original script reuses the
name `ga`), then SGA, MB-SGA-10, MB-SGA-100 — but now GA's `t` has been
tripled (`.0015` $\to$ `.005`) and all three SGA-style runs use `t = 1`
instead of `t = .1`, a ten-fold increase.

**Final coefficient estimates** — now compare these to the *same* true
`beta` as before:

```{r bad-t-estimates, echo=FALSE}
cat("True beta  :", round(t(beta),4), "\n")
cat("GA         :", round(ga2$est,4), "\n")
cat("SGA (b=1)  :", round(b1b$est,4), "\n")
cat("MB-SGA-10  :", round(b10b$est,4), "\n")
cat("MB-SGA-100 :", round(b100b$est,4), "\n")
```

Every method now lands **much farther** from the true `beta` than in
Comparison 1 — most obviously GA and MB-SGA-100, whose final estimates are
several times larger in magnitude than the true coefficients (a classic
symptom of a too-large step size: the update overshoots the optimum instead
of approaching it). Correspondingly, the final full-data gradient norms are
much larger than before instead of settling near zero:

```{r bad-t-final-grad, echo=FALSE}
cat("GA         :", round(sum(abs(ga2$grad[ga2$iter,])),2), "\n")
cat("SGA (b=1)  :", round(sum(abs(b1b$grad[b1b$iter,])),2), "\n")
cat("MB-SGA-10  :", round(sum(abs(b10b$grad[b10b$iter,])),2), "\n")
cat("MB-SGA-100 :", round(sum(abs(b100b$grad[b100b$iter,])),2), "\n")
```

Note in particular that GA's final gradient here (over 2,600) is now *much
larger* than any of the stochastic methods' — the opposite of what happened
with well-chosen step sizes. This is the plot's main lesson: a deterministic
method (GA) with too large a step size doesn't just converge slowly, it can
fail to converge at all and oscillate indefinitely, while the *inherent*
per-step noise in SGA/mini-batch can sometimes partially mask (or, as here,
appear less severe in this summary statistic than) the instability, even
though all four have moved well away from the true optimum.

## The Second Comparison Plot

```{r plot-bad-t, fig.width=7, fig.height=5}
index <- 1:1000
plot(apply(ga2$grad[index,], 1, function(t) sum(abs(t))), type = 'l', ylim =
       c(0,max(apply(b1b$grad[,], 1, function(t) sum(abs(t))))), ylab =
       "Complete gradient")
lines(apply(b1b$grad[index,], 1, function(t) sum(abs(t))), col = "red")
lines(apply(b10b$grad[index,], 1, function(t) sum(abs(t))), col = "blue")
lines(apply(b100b$grad[index,], 1, function(t) sum(abs(t))), col = "orange")
legend("topright", col = c("black", "red", "blue", "orange"), legend =
         c("GA", "SGA", "MB-SGA-10", "MB-SGA-100"), lty = 1)
```

This time `index <- 1:1000` plots *all* iterations (not just the first 500),
since the interesting behavior — sustained oscillation — persists
throughout the whole run rather than settling down quickly.

**Reading the plot:** the black GA line is dramatically different from
Comparison 1 — instead of smoothly decreasing to near zero, it **oscillates
with large, persistent swings** for hundreds of iterations, only slowly
damping down (and still ending far above zero, matching the large final
gradient value confirmed above). This is the textbook visual signature of a
step size that is too large for gradient ascent/descent: the update
overshoots the optimum, lands on the "other side," overshoots again in the
opposite direction, and so on — sometimes damping out slowly, sometimes not
damping at all. The SGA and mini-batch curves (red, blue, orange) are
noisier than in Comparison 1 too, but their smaller effective step
sizes per observation make the instability less visually dramatic than
GA's — even though, as the coefficient estimates above confirm, none of the
four methods are actually finding the true MLE in this comparison.

# Summary

| Concept | What it does |
|---|---|
| `f.gradient()` | Computes the (rescaled by `n`) logistic-regression score $X^\top(y-\pi)$, usable on either the full data or a mini-batch |
| `SGA()` | One general function: `batch.size = n` gives GA, `batch.size = 1` gives SGA, in-between gives mini-batch SGA |
| Mini-batching | Data is shuffled once, split into `K` fixed batches, and cycled through repeatedly (not reshuffled each epoch) |
| Update rule | $\beta \leftarrow \beta + t_k \cdot (\text{mini-batch gradient})/\text{batch.size}$ |
| `mean_beta` | Running average of iterates — used as the reported estimate for SGA/mini-batch (not for full-batch GA) to smooth out noise |
| `track.gradient` | At every iteration, records the *true full-data* gradient at the current estimate, purely for later comparison/plotting |
| Comparison 1 (good `t`) | All four methods converge close to the true simulated `beta`; GA reaches gradient $\approx 0$ exactly, SGA/mini-batch settle near but not at zero |
| Comparison 2 (bad, large `t`) | All four methods end up far from the true `beta`; GA visibly oscillates for hundreds of iterations without damping to zero |

The script is a compact, hands-on demonstration of the classic
speed-vs-noise trade-off between full-batch gradient ascent and its
stochastic/mini-batch relatives, and of how an improperly large step size
can break convergence for all of them — most dramatically for the
otherwise-well-behaved, deterministic full-batch method.

# Reproducibility Note

This document's narrative numbers (the true simulated `beta`, all four
final coefficient estimates in both comparisons, and all final gradient
values) were obtained by actually executing this script end-to-end in R.
Because `set.seed(10)` fixes the random number generator before any random
data is generated, knitting this `.Rmd` file yourself will regenerate
*exactly* the same simulated `X`, `beta`, and `y`, and therefore exactly the
same downstream results shown above — no packages beyond base R are
required to run this script.
