---
title: "Simulated Annealing Demonstrative Example"
author: "SULAGNA ROY"
date: "2026-09-12"
output: pdf_document
---
#####################################
# Simulated Annealing
# Demonstrative example
#####################################

## This R code demonstrates Simulated Annealing for finding the value x that maximizes
## f(x)=(cos(50*x)+sin(20*x))^2 ,0<x<1.

# Step 1 : Objective Function
```{r}
set.seed(1)
fn <- function(x, T = 1){
  h <- ( cos(50*x) + sin(20*x) )^2 # The actual function being optimized 
  exp(h/T) * (0 < x & x < 1) # ensures that values outside (0,1) receive 0
  # but the algorithm uses exp(h/T) as the target/acceptance function.
}
```


# Step 2 : Plot of Objective Function
```{r}
x <- seq(0, 1, length = 5e2)
plot(x, fn(x), type = 'l', ylim = c(0,250), ylab = "exp(f/T)",lwd=1.5)

# Effect of Temperature
tseq = c(.83, .75, .71)

# The temperature T controls how strongly the algorithm concentrates around high values of h(x).

# High T implies flatter distribution implying more exploration
# Low T implies sharper peaks implying more exploitation

# So as T changes from 0.83 to 0.75 to 0.71 the peaks become increasingly pronounced.

for(t in 1:3){
  lines(x, fn(x, T = tseq[t]), col = t+1,lwd=1.5)
}
legend("topright", col = 1:4, lty = 1, 
       legend = c("T = 1", "T = .83", "T = .75", "T = .71"))

```

# Step 3: Simulated Annealing Algorithm
```{r}
# r=.3 controls the size of the possible movement
simAn <- function(N = 10, r = .3){
  x <- numeric(length = N)
  
  # Initial Value
  x[1] <- runif(1) 
  
  for(k in 2:N){
    
    # Generate a candidate
    a <- runif(1, x[k-1] - r,  x[k-1] + r)
    
    # Decrease Temperature this is the cooling schedule
    T <- 1/(log(k))
    
    # Calculate Acceptance Ratio
    ratio <- fn(a,T)/fn(x[k-1], T)  #fn is the exp(h/T)
    if( runif(1) < ratio)
    {
       # If the ratio is greater than 1 candidate is automatically accepted
      x[k] <- a
    }else
    {
      x[k] <- x[k-1] 
      # If the candidate is worse, the ratio is less than 1, 
      # but it can still be accepted with some probability.
      # This is the key idea of simulated annealing: 
      # occasionally accepting a worse solution helps avoid getting trapped in a local maximum.
    }
  }
 return(x) 
}
```


# Step 4: Find the estimated Optimizer
```{r}
N <- 500
sim <- simAn(N = N)

# This finds the simulated value having the largest value of fn()
sim[which.max(fn(sim))]  # theta^*
```

# Step 5: Final Plot
```{r}
plot(x, fn(x), type = 'l', ylab = "exp(f/T)")
points(sim, fn(sim), pch = 16, col = 1)
points(sim[which.max(fn(sim))],fn(sim[which.max(fn(sim))]),pch=17,col="red")
```

# Step 6: Comment
Simulated annealing starts by exploring the whole parameter space and gradually 
becomes more selective as the temperature decreases, 
allowing it to escape local maxima and approach the global maximum.

# ---------------------------------------------------------------------------

