---
title: 'Cauchy MLE using Simulated Annealing: Complete Code Explanation'
author: "Sulagna Roy"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    highlight: tango
  pdf_document:
    toc: true
    number_sections: true
    keep_tex: false
  word_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
set.seed(1)
```

---

# Introduction

This document provides a comprehensive explanation of R code that implements **Simulated Annealing** to find the Maximum Likelihood Estimate (MLE) of the location parameter μ for a Cauchy distribution.

**Key Challenge:** The log-likelihood for Cauchy distribution is **multimodal (non-concave)**, making standard optimization methods fail. Simulated Annealing overcomes this by using temperature-dependent probabilistic moves.

---

# Part 1: Initialization and Data Generation

## Code Explanation

```{r data_generation}
set.seed(1)           # Set seed for reproducibility
mu.star <- 5          # True location parameter
n <- 4                # Sample size
X <- rt(n, df = 1) + mu.star
```

## Detailed Explanation

- **`set.seed(1)`**: Ensures reproducible random number generation
- **`mu.star <- 5`**: The true location parameter (what we're trying to estimate)
- **`n <- 4`**: Sample size
- **`rt(n, df = 1)`**: Generates n=4 samples from Student's t-distribution with 1 degree of freedom
  - This is equivalent to the **Cauchy distribution**
  - A heavy-tailed distribution with undefined mean and variance
- **`X <- rt(n, df = 1) + mu.star`**: Shifts the Cauchy samples by adding μ*=5
  - Result: X_i = Z_i + 5, where Z_i ~ Cauchy(0,1)
  - Final data: X ~ Cauchy(μ=5)

## Purpose

This creates sample data from a known distribution so we can:
1. Test whether the algorithm recovers the true parameter (μ*=5)
2. Visualize the likelihood function
3. Demonstrate algorithm convergence

---

# Part 2: Log-Likelihood Function

## Code

```{r log_likelihood_function}
log.like <- function(mu, X, T = 1)
{
  n <- length(X)
  rtn <- -n*log(pi) - sum( log(1 + (X - mu)^2) )
  return(exp(rtn/T))
}
```

## Mathematical Background

The log-likelihood for Cauchy distribution is:

$$\ell(\mu) = -n \ln(\pi) - \sum_{i=1}^{n} \ln(1 + (X_i - \mu)^2)$$

## Code Explanation

- **Input Parameters:**
  - `mu`: Location parameter (what we're optimizing)
  - `X`: Observed data
  - `T`: Temperature (default = 1)

- **`n <- length(X)`**: Gets sample size

- **`rtn <- -n*log(pi) - sum(log(1 + (X - mu)^2))`**: 
  - Calculates the log-likelihood
  - `-n*log(pi)`: First term (constant across μ values)
  - `sum(log(1 + (X - mu)^2))`: Sum of log terms

- **`return(exp(rtn/T))`**: 
  - Returns **transformed likelihood** = exp(ℓ(μ)/T)
  - NOT the standard likelihood

## Why exp(ℓ/T)?

This is the key innovation for optimization:

**Temperature Effect on Landscape:**

- **High T (e.g., T=1):** 
  - exp(ℓ/T) ≈ 1 + ℓ/T
  - Function is smooth and nearly flat
  - Differences between peaks are dampened
  - Algorithm explores broadly (exploitation minimized)

- **Low T (e.g., T=0.1):**
  - exp(ℓ/T) amplifies peaks
  - Sharp, pronounced peaks
  - Differences between solutions are magnified
  - Algorithm refines locally (exploration reduced)

**This creates the exploration → exploitation transition!**

---

# Part 3: Likelihood Visualization

## Code

```{r likelihood_visualization}
mu.x <- seq(-10, 40, length = 1e3)  # 1000 values from -10 to 40
ll.est <- log(sapply(mu.x, log.like, X))  # Evaluate log-likelihood
plot(mu.x, ll.est, type = 'l', 
     ylab = "log-likelihood", 
     xlab = expression(mu))
```

## Step-by-Step Explanation

### Step 1: Create Parameter Grid
```{r eval=FALSE}
mu.x <- seq(-10, 40, length = 1e3)
```

- Creates 1000 equally-spaced values from -10 to 40
- These represent candidate parameter values

### Step 2: Evaluate Likelihood at Each Point
```{r eval=FALSE}
ll.est <- log(sapply(mu.x, log.like, X))
```

- **`sapply(mu.x, log.like, X)`**: 
  - Apply function `log.like` to each value in `mu.x`
  - Returns a vector of transformed likelihoods (with T=1)
  
- **`log(...)`**: 
  - Takes natural logarithm
  - Converts exp(ℓ/T) back to ℓ scale for visualization
  - Makes it easier to see the shape

### Step 3: Visualize

```{r eval=FALSE}
plot(mu.x, ll.est, type = 'l', ylab = "log-likelihood", xlab = expression(mu))
```

- Creates line plot of log-likelihood vs. μ
- Shows the **multimodal structure**

## Key Observation

The log-likelihood has **multiple peaks** (local maxima), which is why:
- **Gradient-based methods fail**: They get trapped in nearest local peak
- **Simulated Annealing is needed**: To jump between peaks and find global maximum

---

# Part 4: Simulated Annealing Algorithm

This is the core algorithm. We'll break it into parts.

## Complete Algorithm

```{r simulated_annealing_function}
simAn <- function(N = 10, r = .5){
  # ========== INITIALIZATION ==========
  x <- numeric(length = N)                    # Store parameter trajectory
  x[1] <- runif(1, min = -10, max = 40)     # Random starting point
  fn.value <- numeric(length = N)            # Store likelihood values
  
  fn.value[1] <- log.like(mu = x[1], X, T = 1)
  
  # ========== MAIN LOOP ==========
  for(k in 2:N){
    
    # STEP 1: Generate Candidate (Random Walk)
    a <- runif(1, x[k-1] - r, x[k-1] + r)
    
    # STEP 2: Temperature Schedule (Cooling)
    T <- 1/(1 + log(log(k)))
    
    # STEP 3: Acceptance Ratio (Metropolis)
    ratio <- log.like(mu = a, X, T)/log.like(mu = x[k-1], X, T)
    
    # STEP 4: Accept or Reject (Metropolis Criterion)
    if( runif(1) < ratio){
      x[k] <- a
    } else{
      x[k] <- x[k-1]
    }
    
    # Store function value at T=1 (for comparison)
    fn.value[k] <- log.like(mu = x[k], X, T = 1)
  }
  
  # Return both trajectory and function values
  return(list("x" = x, "fn.value" = fn.value)) 
}
```

## Detailed Explanation

### Initialization Phase

```{r eval=FALSE}
x <- numeric(length = N)
x[1] <- runif(1, min = -10, max = 40)
fn.value <- numeric(length = N)
fn.value[1] <- log.like(mu = x[1], X, T = 1)
```

- **`x <- numeric(length = N)`**: 
  - Vector to store all N parameter values (the algorithm's trajectory)
  
- **`x[1] <- runif(1, min = -10, max = 40)`**: 
  - Initialize with random value in [-10, 40]
  - Ensures we don't start at a known optimum
  - Allows multiple runs to explore different regions
  
- **`fn.value`**: 
  - Stores likelihood values
  - Always evaluated at T=1 for comparability

---

### Main Loop: Step 1 - Candidate Generation

```{r eval=FALSE}
a <- runif(1, x[k-1] - r, x[k-1] + r)
```

**What it does:**
- Generates a candidate solution `a`
- From uniform distribution centered at current position x_{k-1}
- Range: [x_{k-1} - r, x_{k-1} + r]

**Parameters:**
- **r = 0.5** (default): Step size/neighborhood radius
  - r=0.5 means candidate is within ±0.5 of current position
  - Larger r: Bigger jumps, more exploration, fewer acceptances
  - Smaller r: Smaller jumps, slower exploration, more acceptances

**Purpose:** Implement random walk to explore parameter space locally

---

### Main Loop: Step 2 - Temperature Schedule

```{r eval=FALSE}
T <- 1/(1 + log(log(k)))
```

**Temperature Calculation:**
- Formula: $T_k = \frac{1}{1 + \ln(\ln(k))}$
- Decreases monotonically as k increases (cooling schedule)

**Temperature Evolution:**

```{r temperature_schedule}
k_values <- c(2, 5, 10, 50, 100, 500, 1000)
T_values <- 1/(1 + log(log(k_values)))
schedule_df <- data.frame(Iteration = k_values, Temperature = round(T_values, 4))
print(schedule_df)
```

**Key Properties:**
- **Early iterations** (small k): ln(ln(k)) is small → T is large
  - High temperature = smooth landscape
  - Algorithm accepts bad moves frequently
  - **Broad exploration of parameter space**

- **Late iterations** (large k): ln(ln(k)) is large → T is small
  - Low temperature = sharp peaks
  - Algorithm rejects bad moves frequently
  - **Local refinement around best solution**

- **Asymptotic behavior:** As k → ∞, T → 0
  - Algorithm converges (stops accepting bad moves)

**Why this schedule?**
- Sub-geometric cooling (slower than exponential)
- Theoretical guarantee: Convergence to global optimum as N → ∞
- Faster than geometric but slower than linear cooling

---

### Main Loop: Step 3 - Acceptance Ratio

```{r eval=FALSE}
ratio <- log.like(mu = a, X, T)/log.like(mu = x[k-1], X, T)
```

**What it does:**
$$\alpha = \frac{\ell(a)}{\ell(x_{k-1})} = \exp\left(\frac{\ell(a) - \ell(x_{k-1})}{T}\right)$$

**Interpretation:**
- **If ℓ(a) > ℓ(x_{k-1})**: 
  - Candidate is better (higher likelihood)
  - α > 1
  - Will be accepted (unless u > 1, impossible)
  
- **If ℓ(a) < ℓ(x_{k-1})**: 
  - Candidate is worse (lower likelihood)
  - α < 1
  - Accepted with probability α

**Temperature Effect:**

Suppose ℓ(a) - ℓ(x_{k-1}) = -1 (worse by 1 unit):

- High T (T=0.5): α = exp(-1/0.5) = exp(-2) ≈ 0.135 → Accept with 13.5% probability
- Low T (T=0.1): α = exp(-1/0.1) = exp(-10) ≈ 0.00005 → Accept with 0.005% probability

**Interpretation:** High temperature is more **forgiving** of bad moves (exploration), while low temperature is more **selective** (exploitation).

---

### Main Loop: Step 4 - Metropolis Criterion (Accept/Reject)

```{r eval=FALSE}
if( runif(1) < ratio){
  x[k] <- a
} else{
  x[k] <- x[k-1]
}
```

**Algorithm:**
1. Generate u ~ Uniform(0,1)
2. **If u < α**: Accept candidate (x_k ← a)
3. **Else**: Reject candidate, keep current (x_k ← x_{k-1})

**Probabilistic Acceptance:**
- Since u is uniform on [0,1], P(u < α) = α
- Better solutions (α > 1): Always accepted
- Worse solutions (α ≤ 1): Accepted with probability α

**Why This Works:**

The key insight is that **accepting some bad moves allows escape from local maxima**:

```
START at local maximum → Temperature is high → Accept bad move anyway
       ↓
Move to different region → Climb different peak
       ↓
Temperature decreases → Stop accepting bad moves
       ↓
Converge to global maximum (if lucky) or local maximum (if unlucky)
```

---

### Store Function Value

```{r eval=FALSE}
fn.value[k] <- log.like(mu = x[k], X, T = 1)
```

- Always evaluates at **T=1** (not temperature-adjusted)
- This is the true likelihood (not the transformed version)
- Allows fair comparison across all iterations
- Used later for finding the best solution

---

### Return Statement

```{r eval=FALSE}
return(list("x" = x, "fn.value" = fn.value))
```

Returns a list with:
- **x**: All parameter values sampled (the trajectory)
- **fn.value**: Corresponding likelihood values

---

# Part 5: Testing the Algorithm

## Run Single Algorithm

```{r single_run}
# Run simulated annealing for 100 iterations with r=5
sim <- simAn(N = 1e2, r = 5)

# Find the parameter value with highest likelihood
mu_hat <- sim$x[which.max(sim$fn.value)]
cat("Estimated MLE: mu_hat =", round(mu_hat, 4), "\n")
cat("True parameter: mu* =", mu.star, "\n")
```

## Multiple Independent Runs

```{r multiple_runs, fig.width=12, fig.height=10}
# Create grid for plotting likelihood
mu.x <- seq(-10, 40, length = 1e3)
ll.est <- log(sapply(mu.x, log.like, X))

# Run algorithm 4 times independently
par(mfrow = c(2,2))

colors <- c("blue", "darkred", "darkgreen", "purple")
mles <- numeric(4)

for(i in 1:4){
  sim <- simAn(N = 1e2, r = 5)
  mles[i] <- sim$x[which.max(sim$fn.value)]
  
  plot(mu.x, ll.est, type = 'l', 
       ylab = "log-likelihood", 
       xlab = expression(mu),
       main = paste("Run", i))
  
  points(sim$x, log(sim$fn.value), 
         pch = 16, 
         col = adjustcolor(colors[i], alpha = .4))
}

par(mfrow = c(1,1))

cat("\nMLE estimates from 4 independent runs:\n")
print(round(mles, 4))
cat("Mean of estimates:", round(mean(mles), 4), "\n")
cat("SD of estimates:", round(sd(mles), 4), "\n")
```

## What This Shows

- **Black curve**: True log-likelihood ℓ(μ)
- **Colored dots**: Sampled parameter values during algorithm execution
- **All 4 plots**: Same likelihood curve but different random trajectories
- **Key observation**: All runs converge to same peak despite different starting points

This demonstrates **global optimization** (not trapped in local maxima).

---

# Part 6: Effect of Step Size r

## Large Step Size: r = 500

```{r large_r, fig.width=12, fig.height=5}
par(mfrow = c(1,2))

# Large r = 500
sim <- simAn(N = 1e3, r = 500)
plot(mu.x, ll.est, type = 'l', 
     main = "r = 500. Many rejections",
     ylab = "log-likelihood", 
     xlab = expression(mu))
points(sim$x, log(sim$fn.value), 
       pch = 16, 
       col = adjustcolor("blue", alpha = .2))
cat("\nLarge r = 500:\n")
cat("Number of accepted moves:", sum(diff(sim$x) != 0), "out of 999\n")
cat("Acceptance rate:", round(100*sum(diff(sim$x) != 0)/999, 1), "%\n")
```

**Explanation:**
- **r = 500**: Candidates are ±500 from current position
- **Problem**: Parameter space is only [-10, 40], so proposals often land far outside data likelihood mass
- **Result**: High rejection rate (wasted iterations)
- **Trajectory**: Sparse (few accepted moves, many repeated values)
- **Efficiency**: Poor

---

## Small Step Size: r = 0.1

```{r small_r, fig.width=12, fig.height=5}
# Small r = 0.1
sim <- simAn(N = 1e3, r = 0.1)
plot(mu.x, ll.est, type = 'l', 
     main = "r = 0.1. Many small acceptances",
     ylab = "log-likelihood", 
     xlab = expression(mu))
points(sim$x, log(sim$fn.value), 
       pch = 16, 
       col = adjustcolor("blue", alpha = .2))
cat("\nSmall r = 0.1:\n")
cat("Number of accepted moves:", sum(diff(sim$x) != 0), "out of 999\n")
cat("Acceptance rate:", round(100*sum(diff(sim$x) != 0)/999, 1), "%\n")

par(mfrow = c(1,1))
```

**Explanation:**
- **r = 0.1**: Candidates are ±0.1 from current position
- **Advantage**: Nearly all proposals are accepted (high acceptance rate)
- **Disadvantage**: Algorithm moves very slowly through parameter space
- **Risk**: May not explore broadly enough to find global maximum
- **Trajectory**: Dense (many different sampled values)

---

## Optimal Step Size

**General principle:** 
$$r_{opt} \approx \frac{\text{Parameter Range}}{10}$$

In this case:
- Parameter range: [-10, 40] → width = 50
- Optimal r ≈ 5
- This is why r=5 was used in earlier examples

---

# Summary: Key Concepts

## 1. Temperature Schedule: $T_k = \frac{1}{1 + \ln(\ln(k))}$

- **Decreases over iterations**: Controls landscape shape
- **Early iterations** (high T): Smooth landscape → Broad exploration
- **Late iterations** (low T): Sharp peaks → Local refinement
- **Purpose**: Escape from local maxima early, then converge

## 2. Metropolis Criterion: $\alpha = e^{(\ell(a) - \ell(x_{k-1}))/T}$

- **Always accept** if α > 1 (uphill moves)
- **Probabilistically accept** if α ≤ 1 (downhill moves with probability α)
- **Temperature modulates tolerance**: High T accepts more bad moves
- **Key insight**: Occasional bad moves enable global optimization

## 3. Step Size r

- **Too large**: Rejections dominate, inefficient
- **Too small**: Slow exploration, may miss global maximum
- **Optimal**: r ≈ parameter range / 10
- **Trade-off**: Balance between exploration breadth and acceptance rate

## 4. Convergence Properties

- **Global optimization**: Finds true maximum, not local optimum
- **Robust**: Works from any starting point
- **Multimodal**: Handles non-concave likelihood functions
- **Theoretical guarantee**: With sub-geometric cooling, converges to global optimum as N → ∞

## 5. Why Simulated Annealing Works for Cauchy

- **Problem**: Cauchy likelihood is multimodal (non-concave)
- **Gradient methods fail**: Climb nearest peak, get stuck
- **Simulated Annealing succeeds**: Probabilistically jumps between peaks
- **Cost**: Slower than local methods, but more robust

---

# Conclusion

Simulated Annealing is a powerful algorithm for **global optimization** that:

1. **Mimics physics**: Models cooling of metals (annealing)
2. **Uses temperature**: Controls exploration vs. exploitation
3. **Employs randomness**: Escapes local optima probabilistically
4. **Guarantees convergence**: Under mild conditions, finds global optimum
5. **Applies broadly**: Works for any multimodal optimization problem

The algorithm is particularly valuable for:
- Maximum likelihood estimation in complex models
- Parameter inference with non-concave likelihood
- Any discrete or continuous optimization problem
- Problems where gradients are unavailable or unreliable

---

# References

- Metropolis, N., Rosenbluth, A. W., Rosenbluth, M. N., Teller, A. H., & Teller, E. (1953). Equation of state calculations by fast computing machines. Journal of Chemical Physics.

- Kirkpatrick, S., Gelatt, C. D., & Vecchi, M. P. (1983). Optimization by simulated annealing. Science, 220(4598), 671-680.

